An Optimized Agent for Capture The Flag Pac-Man

### 2011-03-09 projectWriteup/paper.tex
some revisions to writeup

\begin{document}
\maketitle
\section{Introduction}
A popular and relevant field in artificial intelligence is noisy gameplay
\cite{miik}.  Noisy readouts, opposing agents, and limited calculation
time, mean that noisy games are open to a wide variety of AI techniques.
One particularly relevant example of this field is Capture the Flag (CTF)
Pac-Man, in which a team of independent agents tries to simultaneously
protect food on their side of a Pac-Man board while eating the food on
their opponents' side.  In this paper, we discuss our implementation of a
balanced agent to play CTF Pac-Man.
\\ \\
In our balanced agent, we observed that there are three main branches of
strategy in CTF Pac-Man: attacking, holding the line, and defending.  The
first attempts to eat opposing food, the second attempts to block opposing
agents from entering our side, and the third attempts to capture opposing
agents and prevent them from eating our food once they cross over.  We
built individual modules for each one of these strategies to recommend
moves.  As in many AI agents, the recommendations of these modules and the
global decision of which module to apply use a combination of minimax and
expectimax, where we look several moves ahead, analyze the strength of our
position, and act accordingly.
\\ \\
Our differentiator and the primary focus of our work was on more advanced
ways to model opponents and determine the value of a given game state,
as well as techniques to optimize the weights used to combine those
parameters into relevant state values.  For inference of opponent agent
positions, using prior knowledge and the noisy readings we could observe,
we used more than just sensor data, and incorporated a basic opponent
model.  Our various agents incorporated a wide variety of heuristics,
weighted to recommend moves.  We used a technique called Particle Swarm
Optimization (PSO) to find a string set of weights.  PSO tends to perform
well in functions like ours that are difficult to evaluate accurately and
impossible to analyze from a theoretical perspective \cite{carlisle}.
Furthermore, we observed that the best weightings varied between different
layouts.  Therefore, we trained weights for many different boards.  When
presented with a new board, we clustered it with similar boards (based on
a number of features of the layout) for which
we had trained values and used an average of those values.  The result is
an agent that has learned how to play well on a variety of different
layouts with a variety of different strategies, and combines that
knowledge effectively when presented with a new game.

\section{Heuristics}
We thought much more about our choice of heuristics than about their weights, relying instead on automated techniques to choose optimal parameters.  Our choice of heuristics was primarily driven by attempting to think about things in ways similar to humans; in each ``mode" we evaluated the position using heuristics relevant to examining the board from the perspective of the human strategy.

Hence, we wrote a heuristic function that looked at a board and classified each of the agents, theirs and ours, as being in one of those three states.  To evaluate a position we took a weighted sum of how the three heuristics thought we were doing; thus our heuristic was considering a weighting of how well our defenders were protecting our dots, our attackers were threatening their dots and our hold the line agents were preventing them from crossing onto our side.
\subsection{Offense}
\subsection{Defense}
The defensive heuristics that we found most helpful were putting emphasis on staying close to our enemies, getting them into one exit scenarios and ensuring that for any piece of food we were closer than our opponents.  A key limitation that we had was that losing sight of attacking enemies could be very dangerous because lack of information is very helpful to attackers.  However, because we were able to deduce an exact location any time food went missing, any agent that was doing serious harm was inherently easy to track.
\subsection{Hold the Line}
Playing defense against agents who had not crossed onto our side was a very different problem than playing it against agents who had, mostly because trying to eat them was no longer a strong strategy.  Hence we hit upon the idea of separating that type of play into a ``Hold The Line" heuristic.  For the purposes of holding the line, our agents would ``match up'' against usually one but sometimes more, if we had some agents attacking, opposing agents and mirror their moves, trying to stay closer to all of the squares on the border than the opponents that that agent was ``covering''.  With good inference, as is possible in a 6 agent game, it could become very hard for any opposing agents to cross the border.

\section{Inference}
For inference of opponent agent positions, we update prior probabilities based not only on our sensor readings, but also based on assumptions in our opponent model, which assumes that the opponent is at all times more likely to be moving towards our food and, therefore, places higher probabilities on positions that get them closer. Additionally, our inference module analyzes our food grid and, whenever one of our pieces of food gets eaten, updates the probabilities of the enemy agent that ate it (whichever agent moved before our agent who' is currently analyzing his sensor readings). We are also careful in our inference module about cases where we eat an opponent, to update the probabilities to reflect that they are starting over at their initial position.

\section{Particle Swarm Optimization}
The fitness function that we were trying to maximize is, of course, how well an agent plays.  However, evaluating the performance of a strategy is tricky because it requires playing many games to determine a win percentage against some training set of agents and evaluating some measure of fitness.  Evaluating a win percentage is quite noisy and playing a single pacman game is computationally very expensive (as much as 50 minutes, although for training purposes we cutdown thinking time to make games around 5 minutes long).  Hence, when evaluating our agents we got very noisy results.

>>>>

\begin{document}
\maketitle
\section{Introduction}
A popular and relevant field in artificial intelligence is noisy gameplay
\cite{miik}.  Noisy readouts, opposing agents, and limited calculation
time mean that noisy games are open to a wide variety of AI techniques.
One particularly relevant example of this field is Capture the Flag (CTF)
Pac-Man, in which a team of independent agents tries to simultaneously
protect food on their side of a Pac-Man board while eating the food on
their opponents' side.  In this paper, we discuss our implementation of a
balanced agent to play CTF Pac-Man.
\\ \\
In our balanced agent, we observed that there are three main branches of
strategy in CTF Pac-Man: attacking, holding the line, and defending.  The
first attempts to eat opposing food, the second attempts to block opposing
agents from entering our side, and the third attempts to capture opposing
agents and prevent them from eating our food once they cross over.  We
built individual modules for each one of these strategies to recommend
moves.  As in many AI agents, the recommendations of these modules and the
global decision of which module to apply use a combination of minimax and
expectimax, where we look several moves ahead, analyze the strength of our
position, and act accordingly.
\\ \\
Our differentiator and the primary focus of our work was on more advanced
ways to model opponents and determine the value of a given game state,
as well as techniques to optimize the weights used to combine those
parameters into relevant state values.  For inference of opponent agent
positions, using prior knowledge and the noisy readings we could observe,
we used more than just sensor data, and incorporated a basic opponent
model.  Our various agents incorporated a wide variety of heuristics,
weighted to recommend moves.  We used a technique called Particle Swarm
Optimization (PSO) to find a string set of weights.  PSO tends to perform
well in functions like ours that are difficult to evaluate accurately and
impossible to analyze from a theoretical perspective \cite{carlisle}.
Furthermore, we observed that the best weightings varied between different
layouts.  Therefore, we trained weights for many different boards.  When
presented with a new board, we clustered it with similar boards (based on
a number of features of the layout) for which
we had trained values and used an average of those values.  The result is
an agent that has learned how to play well on a variety of different
layouts with a variety of different strategies, and combines that
knowledge effectively when presented with a new game.

\section{Heuristics}
We thought much more about our choice of heuristics than about their weights, relying instead on automated techniques to choose optimal parameters.  Our choice of heuristics was primarily driven by attempting to think about things in ways similar to humans; in each "mode" - offense, defense, and hold the line - we evaluated the positions using heuristics relevant to examining the board from the perspective of the human strategy.

Hence, we wrote a heuristic function that looked at a board and classified each of the agents, theirs and ours, as being in one of those three states.  To evaluate a position we took a weighted sum of how the three heuristics thought we were doing; thus our heuristic was considering a weighting of how well our defenders were protecting our dots, our attackers were threatening their dots and our hold the line agents were preventing them from crossing onto our side.
\subsection{Offense}
\subsection{Defense}
The defensive heuristics that we found most helpful were putting emphasis on staying close to our enemies, getting them into exit scenarios and ensuring that for any piece of food we were closer than our opponents.  A key limitation that we had was that losing sight of attacking enemies could be very dangerous because lack of information is very helpful to attackers.  However, because we were able to deduce an exact location any time food went missing, any agent that was doing serious harm was inherently easy to track.
\subsection{Hold the Line}
Playing defense against agents who had not crossed onto our side was a very different problem than playing it against agents who had, mostly because trying to eat them was no longer a strong strategy.  Hence we hit upon the idea of separating that type of play into a ``Hold The Line" heuristic.  For the purposes of holding the line, our agents would ``match up'' against usually one but sometimes more, if we had some agents attacking, opposing agents and mirror their moves, trying to stay closer to all of the squares on the border than the opponents that that agent was ``covering''.  With good inference, as is possible in a 6 agent game, it could become very hard for any opposing agents to cross the border.

\section{Inference}
For inference of opponent agent positions, we update prior probabilities based not only on our sensor readings, but also based on assumptions in our opponent model, which assumes that the opponent is at all times more likely to be moving towards our food and, therefore, places higher probabilities on positions that get them closer. Additionally, our inference module analyzes our food grid and, whenever one of our pieces of food gets eaten, updates the probabilities of the enemy agent that ate it (whichever agent moved before our agent who is currently analyzing his sensor readings). We are also careful in our inference module about cases where we eat an opponent. When this happens, we update the probabilities to reflect that they're starting over at their initial position. And when an opponent is visible (i.e. in our line of sight), we update our probability mass to show that their position is exactly the one at which we see them. Once they're beyond our line of sight, we return to updating prior probabilities the same way we were before.
\section{Particle Swarm Optimization}
The fitness function that we were trying to maximize is, of course, how well an agent plays.  However, evaluating the performance of a strategy is tricky because it requires playing many games to determine a win percentage against some training set of agents and evaluating some measure of fitness.  Evaluating a win percentage is quite noisy and playing a single pacman game is computationally very expensive (as much as 50 minutes, although for training purposes we cutdown thinking time to make games around 5 minutes long).  Hence, when evaluating our agents we got very noisy results.
